### 2016-06-06 Chapters/theory.tex
Refactoring paragraph 3 and changing a lot of file names

\section*{Introduction}

@@

\section*{Results}

\subsection*{Self-replicator model of resource allocation}
\label{sec:model}

@@

\begin{figure}[!h]
%\includegraphics[scale=1]{./Fig/Fig1.eps}
\caption[Self-replicator model of bacterial growth]{
{\bf Self-replicator model of bacterial growth.} External substrates $S$ enter the cell and are transformed into precursors $P$ through the action of the metabolic machinery $M$.
The precursors are used by the gene expression machinery $R$ to make the proteins composing both the metabolic machinery (transporters, enzymes, ...) and the gene expression machinery itself (RNA polymerase, ribosomes, ...).
$\alpha$ ($1-\alpha$) is the mass proportion of precursors converted into $R$ ($M$).
Thick arrows denote reactions and thin, dashed arrows denote catalytic activities.
The rate of synthesis of precursors and the rate of synthesis of proteins from precursors are denoted by $V_M$ and $V_R$, respectively.
}
\label{fig:self_replicator}
\end{figure}

@@

\subsection*{Growth-rate maximization of the self-replicator reproduces bacterial growth laws}
\label{sec:growthlaws}

@@

\begin{figure}[!h]
%\includegraphics[scale=1]{./Fig/Fig2.eps}
\caption[Analysis of self-replicator model of bacterial growth]{\textbf{Analysis of self-replicator model of bacterial growth.}\\
\textit{A:}~Phase-plane analysis of the self-replicator model of Eqs~\ref{eq:pdef} and \ref{eq:rdef}.
The nullclines for $p$ and $r$ are shown as solid and dashed curves, respectively.
Parameter values are $e_M = 3.6$~h\textsuperscript{-1}, $k_R = 3.6$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, $\beta = 0.003$~L~g\textsuperscript{-1}, $\alpha = 0.45$.
\textit{B:}~Dependence of the growth rate at steady state $\mu^{*}$ on the resource allocation parameter $\alpha$, for two different environmental conditions (solid line, $e_M = 4.76$~h\textsuperscript{-1}; dashed line, $e_M = 1.57$~h\textsuperscript{-1}, other parameter values are $k_R = 2.23$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, and $\beta = 0.003$~L~g\textsuperscript{-1}).
The maximal growth rate is attained for a unique $\alpha$, called $\alpha^*_{opt}$.
}
\label{fig:model_analysis}
\end{figure}

@@

\begin{figure}[!h]
%\includegraphics[scale=1]{./Fig/Fig3.eps}
\caption[Self-replicator model accounts for bacterial growth laws]{
\textbf{Self-replicator model accounts for bacterial growth laws.}\\
\textit{A:}~Predicted quasi-linear relation between the maximal growth rate $\mu^{*}_{opt}$ and the corresponding optimal resource allocation $\alpha^*_{opt}$, for different values of $e_M$ (different colors).
The colored dots indicate $\alpha^*_{opt}$ and $\mu^{*}_{opt}$  for $k_R = 2.23$~h\textsuperscript{-1} and different $e_M$, and the dashed black line the relation for all intermediate values of $e_M$.
The dashed colored lines indicate the relation between $\alpha^*_{opt}$ and $\mu^{*}_{opt}$ obtained when, for a given value of $e_M$, the value of $k_R$ is decreased (lower $k_R$ leads to lower $\mu^{*}_{opt}$).
The solid grey curves correspond to $(\mu^*,\alpha)$-profiles like those shown in Fig.~\ref{fig:model_analysis}\textit{B}.
\textit{B:}~Measured relation between the total RNA/protein mass ratio and the growth rate, in different growth media with different doses of a translation inhibitor (data from~\cite{scott_interdependence_2010}).
For each medium, indicated by a color, five different concentrations of inhibitor were used (higher dose leads to lower growth rate).
Growth-medium compositions are given in the original publication and error bars represent standard deviations.
The dashed black and colored lines are the same as in panel \textit{A}, indicating the good quantitative correspondence between model predictions and experimental data for the chosen parameter values, obtained by fitting the model to the data points (see \textit{Methods} for details).
}
\label{fig:model_validation}
\end{figure}

@@

\subsection*{Biomass maximization as an optimal control problem}
\label{sec:optimalcontrolproblem}

@@

\subsection*{Solution of the optimal control problem}
\label{sec:optimalcontrolsolution}

@@

\begin{figure}[!h]
\centering
%\includegraphics[scale=1]{./Fig/Fig4.eps}
\caption[Optimal control of the self-replicator during a nutrient upshift]{
{\bf Optimal control of the self-replicator during a nutrient upshift.}\\
\textit{A:}~Optimal trajectory in the phase plane for the nondimensionalized model of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, with streamlines. The optimal trajectory is shown as a solid, red curve.
The solid, black curve represents the $\hat{p}$-nullcline.
The dashed, black curve is the switching curve $\varphi(\hat{p})$.
The optimal solution was obtained by numerical optimization using \texttt{bocop}~\cite{bonnans_bocop_2012} (see \textit{Methods} for details), using the parameter values $E_M=1$ and $K=0.003$, and starting from the initial state $(0.024,0.18)$ at $t=0$ (optimal steady state for $E_M=0.2$).
\textit{B:}~Time evolution of the control variable $\alpha_{opt}(\cdot)$ (thick, red line) and the environment $E_M$ (dashed, black line).
}
\label{fig:optimalcontrol}
\end{figure}

@@

\subsection*{Simple feedback control strategies: exploiting information on nutrients or precursors}
\label{sec:steadystate}

@@

\begin{figure}[!h]
\centering
%\includegraphics[scale=1]{./Fig/Fig5.eps}
\caption[Alternative strategies for controlling the self-replicator of bacterial growth]{
{\bf Alternative strategies for controlling the self-replicator of bacterial growth.}\\
The feedback control strategies, shown in red and superposed on the self-replicator of Fig.~\ref{fig:self_replicator}, exploit information on system variables and the environment to adjust the value of $\alpha$, and thus the relative allocation of resources to the metabolic machinery and gene expression machinery.
}
\label{fig:strategies_overview}
\end{figure}

@@

\begin{figure}[!h]
\centering
%\includegraphics[scale=1]{./Fig/Fig6.eps}
\caption[Comparison of the performance of the nutrient-only and precursor-only strategies after a nutrient upshift]{
{\bf Comparison of the performance of the nutrient-only and precursor-only strategies after a nutrient upshift.}\\
\textit{A:}~Trajectory in the phase plane for the nutrient-only strategy (green curve).
The solid, black curve represents the $\hat{p}$-nullcline.
The dashed, black curve is the $\hat{r}$-nullcline.
The solution is obtained by numerical simulation of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, supplemented with $\alpha = f(E_M)$ as specified by Eq.~\ref{eq:meth_f} in the \textit{Methods} section and plotted in \ref{S1_Fig}.
The initial state corresponds to the steady state attained for an environment given by $0.2 \, E_M$.
While converging to the new steady state after the upshift, the precursor concentration makes a large overshoot.
\textit{B:}~As above, but for the precursor-only strategy.
The feedback control strategy is now defined by $\alpha = g(\hat{p})$ as specified by Eq.~\ref{eq:meth_g} in the \textit{Methods} section and plotted in \ref{S1_Fig}.
The solution trajectory (blue curve) exhibits a lower overshoot.
\textit{C:}~Evolution of the control variable $\alpha(\cdot)$ as a function of time, for each of the above two strategies.
Notice that in the nutrient-only strategy $\alpha(\cdot)$ immediately jumps to the optimal value for the post-upshift steady state (green curve), whereas in the precursor-only strategy it depends on the (time-varying) precursor concentration (blue curve).
\textit{D:}~Evolution of the ratio $\texttt{Vol} / \texttt{Vol}_{opt}$  as a function of time, where $\texttt{Vol}$ is the volume of the self-replicator and $\texttt{Vol}_{opt}$ the volume of the same replicator following the optimal strategy shown in Fig.~\ref{fig:optimalcontrol}.
In all of the above simulations, the parameter values $E_M=1$ and $K=0.003$ were used.
}
\label{fig:comparison_nutrients_precursor}
\end{figure}

@@

\subsection*{A near-optimal feedback control strategy: exploiting information on the imbalance between precursors and the gene expression machinery}
\label{sec:strategies}

@@

\begin{figure}[!h]
\centering
%\includegraphics[scale=1]{./Fig/Fig7.eps}
\caption[Comparison of the performance of the precursor-only and the on-off strategies after a nutrient upshift.]
{
{\bf Comparison of the performance of the precursor-only and the on-off strategies after a nutrient upshift.}\newline
\textit{A:}~Trajectory in the phase plane for the on-off strategy (yellow curve).
The solid, black curve represents the $\hat{p}$-nullcline and the dashed, black curve the function $g$.
The solution is obtained by numerical simulation of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, supplemented with the equation $\alpha = h(\hat{p},\hat{r})$ defined in Eq.~\ref{eq:stratswitch} and plotted in Fig.~\ref{fig:ppGppsurface}\textit{A}.
The initial state corresponds to the optimal steady state attained for an environment given by $0.2 \, E_M$.
\textit{B:}~Trajectory in the phase plane for the precursor-only strategy (same as in Fig.~\ref{fig:comparison_nutrients_precursor}\textit{B}, added for comparison).
\textit{C:}~Evolution of the control variable $\alpha$ for each strategy as a function of time.
Both strategies stabilize the system at the optimal steady state, but only the on-off strategy (yellow curve) exhibits bang-bang behavior.
\textit{D:}~Evolution of the ratio $\texttt{Vol} / \texttt{Vol}_{opt}$ for the on-off and precursor-only strategies as a function of time, where $\texttt{Vol}$ is the volume of the self-replicator and $\texttt{Vol}_{opt}$ the volume of the same replicator following the optimal strategy shown in Fig.~\ref{fig:optimalcontrol}.
The final values of $\texttt{Vol} / \texttt{Vol}_{opt}$ attained by the two strategies are 0.9831 and 0.9413, respectively.
The on-off strategy is thus hardly distinguishable from the optimal control strategy in the plot.
In all of the above simulations, the parameter values $E_M=1$ and $K=0.003$ were used.
}
\label{fig:onoffresults}
\end{figure}

@@

\begin{figure}[!h]
\centering
%\includegraphics[scale=1]{./Fig/Fig8.eps}
\caption[ppGpp regulation implements a on-off control strategy of resource allocation.]
{
{\bf ppGpp regulation implements an on-off control strategy of resource allocation.}\newline
\textit{A:}~Response surface of the on-off control strategy, defined by $\alpha = h(\hat{p},\hat{r})$ in Eq.~\ref{eq:stratswitch}.
\textit{B:}~Response surface of the ppGpp control strategy, as defined by Eq.~\ref{eq:ppGpp_empirical} and the simplified kinetic model defining ppGpp in terms of the total amino acid concentration and the ribosomal protein fraction (\ref{S4_Text}).
The shape of the response surface of the ppGpp control strategy is seen to be in very good agreement with the on-off strategy leading to near-optimal performance of the self-replicator during a nutrient upshift.
}
\label{fig:ppGppsurface}
\end{figure}

\section*{Discussion}
\label{sec:discussion}

@@

>>>>

\chapter{Dynamical Allocation of Cellular Resources as an Optimal Control Problem: Novel Insights into Microbial Growth Strategies }
\chaptermark{Dynamical allocation of cell resources}

\selectlanguage{french}
\section*{Résumé du chapître \thechapter : [titre français]}
dans cette section blablabla
\selectlanguage{english}

{\Large
\textbf\newline{\textbf{}}
}

\section{Abstract}
Microbial physiology exhibits growth laws that relate the macromolecular composition of the cell to the growth rate.
Recent work has shown that these empirical regularities can be derived from coarse-grained models of resource allocation.
While these studies focus on steady-state growth, such conditions are rarely found in natural habitats, where microorganisms are continually challenged by environmental fluctuations.
The aim of this paper is to extend the study of microbial growth strategies to dynamical environments, using a self-replicator model.
We formulate dynamical growth maximization as an optimal control problem that can be solved using Pontryagin’s Maximum Principle.
We compare this theoretical gold standard with different possible implementations of growth control in bacterial cells.
We find that simple control strategies enabling growth-rate maximization at steady state are suboptimal for transitions from one growth regime to another, for example when shifting bacterial cells to a medium supporting a higher growth rate.
A near-optimal control strategy in dynamical conditions is shown to require information on several, rather than a single physiological variable.
Interestingly, this strategy has structural analogies with the regulation of ribosomal protein synthesis by ppGpp in the enterobacterium \textit{Escherichia coli}.
It involves sensing a mismatch between precursor and ribosome concentrations, as well as the adjustment of ribosome synthesis in a switch-like manner.
Our results show how the capability of regulatory systems to integrate information about several physiological variables is critical for optimizing growth in a changing environment.

%%%%%% CORE OF THE ARTICLE %%%%%
\section{Introduction}

@@

\section{Results}

\section{Self-replicator model of resource allocation}
\label{sec:model}

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{./Fig/Fig1.eps}
\caption[Self-replicator model of bacterial growth]{
{\bf Self-replicator model of bacterial growth.} External substrates $S$ enter the cell and are transformed into precursors $P$ through the action of the metabolic machinery $M$.
The precursors are used by the gene expression machinery $R$ to make the proteins composing both the metabolic machinery (transporters, enzymes, ...) and the gene expression machinery itself (RNA polymerase, ribosomes, ...).
$\alpha$ ($1-\alpha$) is the mass proportion of precursors converted into $R$ ($M$).
Thick arrows denote reactions and thin, dashed arrows denote catalytic activities.
The rate of synthesis of precursors and the rate of synthesis of proteins from precursors are denoted by $V_M$ and $V_R$, respectively.
}
\label{fig:self_replicator}
\end{figure}

@@

\section{Growth-rate maximization of the self-replicator reproduces bacterial growth laws}
\label{sec:growthlaws}

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{./Fig/Fig2.eps}
\caption[Analysis of self-replicator model of bacterial growth]{\textbf{Analysis of self-replicator model of bacterial growth.}\\
\textit{A:}~Phase-plane analysis of the self-replicator model of Eqs~\ref{eq:pdef} and \ref{eq:rdef}.
The nullclines for $p$ and $r$ are shown as solid and dashed curves, respectively.
Parameter values are $e_M = 3.6$~h\textsuperscript{-1}, $k_R = 3.6$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, $\beta = 0.003$~L~g\textsuperscript{-1}, $\alpha = 0.45$.
\textit{B:}~Dependence of the growth rate at steady state $\mu^{*}$ on the resource allocation parameter $\alpha$, for two different environmental conditions (solid line, $e_M = 4.76$~h\textsuperscript{-1}; dashed line, $e_M = 1.57$~h\textsuperscript{-1}, other parameter values are $k_R = 2.23$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, and $\beta = 0.003$~L~g\textsuperscript{-1}).
The maximal growth rate is attained for a unique $\alpha$, called $\alpha^*_{opt}$.
}
\label{fig:model_analysis}
\end{figure}

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{./Fig/Fig3.eps}
\caption[Self-replicator model accounts for bacterial growth laws]{
\textbf{Self-replicator model accounts for bacterial growth laws.}\\
\textit{A:}~Predicted quasi-linear relation between the maximal growth rate $\mu^{*}_{opt}$ and the corresponding optimal resource allocation $\alpha^*_{opt}$, for different values of $e_M$ (different colors).
The colored dots indicate $\alpha^*_{opt}$ and $\mu^{*}_{opt}$  for $k_R = 2.23$~h\textsuperscript{-1} and different $e_M$, and the dashed black line the relation for all intermediate values of $e_M$.
The dashed colored lines indicate the relation between $\alpha^*_{opt}$ and $\mu^{*}_{opt}$ obtained when, for a given value of $e_M$, the value of $k_R$ is decreased (lower $k_R$ leads to lower $\mu^{*}_{opt}$).
The solid grey curves correspond to $(\mu^*,\alpha)$-profiles like those shown in Fig.~\ref{fig:model_analysis}\textit{B}.
\textit{B:}~Measured relation between the total RNA/protein mass ratio and the growth rate, in different growth media with different doses of a translation inhibitor (data from~\cite{scott_interdependence_2010}).
For each medium, indicated by a color, five different concentrations of inhibitor were used (higher dose leads to lower growth rate).
Growth-medium compositions are given in the original publication and error bars represent standard deviations.
The dashed black and colored lines are the same as in panel \textit{A}, indicating the good quantitative correspondence between model predictions and experimental data for the chosen parameter values, obtained by fitting the model to the data points (see \textit{Methods} for details).
}
\label{fig:model_validation}
\end{figure}

@@

\section{Biomass maximization as an optimal control problem}
\label{sec:optimalcontrolproblem}

@@

\section{Solution of the optimal control problem}
\label{sec:optimalcontrolsolution}

@@

\begin{figure}[!h]
\centering
\includegraphics[width=\textwidth]{./Fig/Fig4.eps}
\caption[Optimal control of the self-replicator during a nutrient upshift]{
{\bf Optimal control of the self-replicator during a nutrient upshift.}\\
\textit{A:}~Optimal trajectory in the phase plane for the nondimensionalized model of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, with streamlines. The optimal trajectory is shown as a solid, red curve.
The solid, black curve represents the $\hat{p}$-nullcline.
The dashed, black curve is the switching curve $\varphi(\hat{p})$.
The optimal solution was obtained by numerical optimization using \texttt{bocop}~\cite{bonnans_bocop_2012} (see \textit{Methods} for details), using the parameter values $E_M=1$ and $K=0.003$, and starting from the initial state $(0.024,0.18)$ at $t=0$ (optimal steady state for $E_M=0.2$).
\textit{B:}~Time evolution of the control variable $\alpha_{opt}(\cdot)$ (thick, red line) and the environment $E_M$ (dashed, black line).
}
\label{fig:optimalcontrol}
\end{figure}

@@

\section{Simple feedback control strategies: exploiting information on nutrients or precursors}
\label{sec:steadystate}

@@

\begin{figure}[!h]
\centering
\includegraphics[width=\textwidth]{./Fig/Fig5.eps}
\caption[Alternative strategies for controlling the self-replicator of bacterial growth]{
{\bf Alternative strategies for controlling the self-replicator of bacterial growth.}\\
The feedback control strategies, shown in red and superposed on the self-replicator of Fig.~\ref{fig:self_replicator}, exploit information on system variables and the environment to adjust the value of $\alpha$, and thus the relative allocation of resources to the metabolic machinery and gene expression machinery.
}
\label{fig:strategies_overview}
\end{figure}

@@

\begin{figure}[!h]
\centering
\includegraphics[height=5cm]{./Fig/Fig6.eps}
\caption[Comparison of the performance of the nutrient-only and precursor-only strategies after a nutrient upshift]{
{\bf Comparison of the performance of the nutrient-only and precursor-only strategies after a nutrient upshift.}\\
\textit{A:}~Trajectory in the phase plane for the nutrient-only strategy (green curve).
The solid, black curve represents the $\hat{p}$-nullcline.
The dashed, black curve is the $\hat{r}$-nullcline.
The solution is obtained by numerical simulation of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, supplemented with $\alpha = f(E_M)$ as specified by Eq.~\ref{eq:meth_f} in the \textit{Methods} section and plotted in \ref{S1_Fig}.
The initial state corresponds to the steady state attained for an environment given by $0.2 \, E_M$.
While converging to the new steady state after the upshift, the precursor concentration makes a large overshoot.
\textit{B:}~As above, but for the precursor-only strategy.
The feedback control strategy is now defined by $\alpha = g(\hat{p})$ as specified by Eq.~\ref{eq:meth_g} in the \textit{Methods} section and plotted in \ref{S1_Fig}.
The solution trajectory (blue curve) exhibits a lower overshoot.
\textit{C:}~Evolution of the control variable $\alpha(\cdot)$ as a function of time, for each of the above two strategies.
Notice that in the nutrient-only strategy $\alpha(\cdot)$ immediately jumps to the optimal value for the post-upshift steady state (green curve), whereas in the precursor-only strategy it depends on the (time-varying) precursor concentration (blue curve).
\textit{D:}~Evolution of the ratio $\texttt{Vol} / \texttt{Vol}_{opt}$  as a function of time, where $\texttt{Vol}$ is the volume of the self-replicator and $\texttt{Vol}_{opt}$ the volume of the same replicator following the optimal strategy shown in Fig.~\ref{fig:optimalcontrol}.
In all of the above simulations, the parameter values $E_M=1$ and $K=0.003$ were used.
}
\label{fig:comparison_nutrients_precursor}
\end{figure}

@@

\section{A near-optimal feedback control strategy: exploiting information on the imbalance between precursors and the gene expression machinery}
\label{sec:strategies}

@@

\begin{figure}[!h]
\centering
\includegraphics[height=5cm]{./Fig/Fig7.eps}
\caption[Comparison of the performance of the precursor-only and the on-off strategies after a nutrient upshift.]
{
{\bf Comparison of the performance of the precursor-only and the on-off strategies after a nutrient upshift.}\newline
\textit{A:}~Trajectory in the phase plane for the on-off strategy (yellow curve).
The solid, black curve represents the $\hat{p}$-nullcline and the dashed, black curve the function $g$.
The solution is obtained by numerical simulation of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, supplemented with the equation $\alpha = h(\hat{p},\hat{r})$ defined in Eq.~\ref{eq:stratswitch} and plotted in Fig.~\ref{fig:ppGppsurface}\textit{A}.
The initial state corresponds to the optimal steady state attained for an environment given by $0.2 \, E_M$.
\textit{B:}~Trajectory in the phase plane for the precursor-only strategy (same as in Fig.~\ref{fig:comparison_nutrients_precursor}\textit{B}, added for comparison).
\textit{C:}~Evolution of the control variable $\alpha$ for each strategy as a function of time.
Both strategies stabilize the system at the optimal steady state, but only the on-off strategy (yellow curve) exhibits bang-bang behavior.
\textit{D:}~Evolution of the ratio $\texttt{Vol} / \texttt{Vol}_{opt}$ for the on-off and precursor-only strategies as a function of time, where $\texttt{Vol}$ is the volume of the self-replicator and $\texttt{Vol}_{opt}$ the volume of the same replicator following the optimal strategy shown in Fig.~\ref{fig:optimalcontrol}.
The final values of $\texttt{Vol} / \texttt{Vol}_{opt}$ attained by the two strategies are 0.9831 and 0.9413, respectively.
The on-off strategy is thus hardly distinguishable from the optimal control strategy in the plot.
In all of the above simulations, the parameter values $E_M=1$ and $K=0.003$ were used.
}
\label{fig:onoffresults}
\end{figure}

@@

\begin{figure}[!h]
\centering
\includegraphics[width=\textwidth]{./Fig/Fig8.eps}
\caption[ppGpp regulation implements a on-off control strategy of resource allocation.]
{
{\bf ppGpp regulation implements an on-off control strategy of resource allocation.}\newline
\textit{A:}~Response surface of the on-off control strategy, defined by $\alpha = h(\hat{p},\hat{r})$ in Eq.~\ref{eq:stratswitch}.
\textit{B:}~Response surface of the ppGpp control strategy, as defined by Eq.~\ref{eq:ppGpp_empirical} and the simplified kinetic model defining ppGpp in terms of the total amino acid concentration and the ribosomal protein fraction (\ref{S4_Text}).
The shape of the response surface of the ppGpp control strategy is seen to be in very good agreement with the on-off strategy leading to near-optimal performance of the self-replicator during a nutrient upshift.
}
\label{fig:ppGppsurface}
\end{figure}

\section{Discussion}
\label{sec:discussion}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Methods}

\section{Steady-state analysis of model}
\label{sec:methods_model_analysis}

The nondimensional version of the model, given by Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, was used for a steady-state analysis of the self-replicator.
Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} were derived from the original model of Eqs~\ref{eq:pdef}-\ref{eq:rdef} by means of the following rescalings:
\begin{equation*}
\hat{p}  = \beta \, p,\;\;\;
\hat{r}  = \beta \, r,\;\;\;
\hat{t}  = k_R \, t,\;\;\;
E_M = e_M/k_R,\;\;\;
K = \beta K_R.
\end{equation*}

As shown in \ref{S1_Text}, for a constant environment $E_M$ and constant resource allocation $\alpha$, the system has two steady states: a trivial unstable steady state $(\hat{p}^*, \hat{r}^*) = (0,1)$, allowing no growth in the absence of precursors, and a steady state with a positive growth rate given by
\begin{equation}
\label{eq:meth_steadystate}
(\hat{p}^*, \hat{r}^*) = \left( \frac{(1-\alpha)\, E_M - \alpha + \sqrt{[(1-\alpha)\, E_M - \alpha]^2 + 4\alpha\, (1-\alpha)\, E_M\, K}}{2\alpha}, \alpha \right).
\end{equation}
The two eigenvalues of the Jacobian matrix evaluated at $(\hat{p}^*, \hat{r}^*)$ are negative (\ref{S1_Text}), so that this steady state is stable.

The growth rate at steady state, as a function of $\hat{p}^*$ and $\hat{r}^*$, is given by Eq.~\ref{eq:munondim}, which we repeat here for clarity:
\begin{equation*}
\hat{\mu}^* = \frac{\hat{p}^*}{K + \hat{p}^*} \, \hat{r}^* .
\end{equation*}
Evaluating $d\hat{p}/dt=0$ at $(\hat{p}^*, \hat{r}^*)$ allows $\hat{r}^*$, and therefore $\hat{\mu}^*$, to be written as a function of $\hat{p}^*$ (\ref{S1_Text}).
Accordingly, we can compute $\partial\hat{\mu}^*/\partial\hat{p}^*$ and, when setting this partial derivative to 0, determine the maximum growth rate at steady state $\mu_{opt}^*$ and the optimal resource allocation $\alpha_{opt}^*$ bringing about this maximal growth rate.
As shown in \ref{S1_Text}, $\mu_{opt}^*$ and $\alpha_{opt}^*$ can be written as explicit functions of either the environment $E_M$:
\begin{equation}
\label{eq:alpha_mu_optimal}
\alpha_{opt}^* = \frac{E_M + \sqrt{K\, E_M}}{E_M + 2\sqrt{K\, E_M} + 1}
, \;\;\;\;\;\;\;\;\;\;\;\;\;\; 
\hat{\mu}^*_{opt} = \frac{E_M}{E_M + 2\sqrt{K\, E_M} + 1},
\end{equation}
or the precursor abundance $\hat{p}_{opt}^*$:
\begin{equation}
\label{eq:alpha_mu_optimal_p}
\alpha_{opt}^* = \frac{\hat{p}^{*}_{opt}}{\hat{p}^{*}_{opt} + \frac{K}{K+\hat{p}^{*}_{opt}}(1+\hat{p}^{*}_{opt})}
, \;\;\;\;\;\;\;\;\;\;\;\;\;\; 
\hat{\mu}^*_{opt} = \frac{\hat{p}^{* 2}_{opt}}{\hat{p}^{* 2}_{opt} + 2K\hat{p}^*_{opt} + K}.
\end{equation}
The above equations were used for the derivation of the control strategies (see below).

\section{Model fitting}
\label{sec:methods_model_fitting}

As can be seen by comparing Figs~\ref{fig:model_validation}\textit{A} and~\ref{fig:model_validation}\textit{B}, growth-rate maximization in the self-replicator model leads to a good qualitative correspondence with the growth laws.
In order to determine if a good quantitative fit of the model with the data from Scott~\textit{et al.} \cite{scott_interdependence_2010} can be obtained, for reasonable parameter values, we estimated $e_M$ and $k_R$ in Eqs~\ref{eq:pdef}-\ref{eq:rdef} from the measured RNA/protein mass ratios.
At steady state, the RNA/protein mass ratio can be interpreted as proportional to $\hat{r}^*$ (and thus $\alpha_{opt}^*$), with an unknown (dimensionless) proportionality constant $\gamma$ (see~\cite{scott_interdependence_2010} for details on the use of the RNA/protein mass ratio as a proxy for the ribosomal protein mass fraction):
\begin{equation}
\hat{r}^* = \alpha_{opt}^* = \gamma \, \frac{\text{RNA mass}}{\text{protein mass}}.
\label{eq:gamma}
\end{equation}
Reformulating Eq.~\ref{eq:alpha_mu_optimal} in terms of the original parameters $e_M$ and $k_R$, which have physical dimensions facilitating the biological interpretation of their values, we obtain a straighforward relation between $e_M$, $k_R$, $K$, $\alpha_{opt}^*$ and $\mu_{opt}^*$:
\begin{equation}
\label{eq:alpha_mu_optimal_dim}
\alpha_{opt}^* = \frac{e_M + \sqrt{K\, e_M\, k_R}}{e_M + 2\sqrt{K\, e_M\, k_R} + k_R}
, \;\;\;\;\;\;\;\;\;\;\;\;\;\; 
\mu^*_{opt} = \frac{e_M \, k_R}{e_M + 2\sqrt{K\, e_M\, k_R} + k_R}.
\end{equation}

Eqs~\ref{eq:gamma}-\ref{eq:alpha_mu_optimal_dim} were used to estimate values of $k_R$ and $\gamma$, as well as $e_M$ for each of the six growth conditions, from the measurements of the growth rate and the RNA/protein mass ratio. 
The value $K$ was not estimated from the experimental data, but set to a value inferred from the literature (\ref{S1_Text}).
The optimization process was carried out by means of the differential evolution algorithm of Storn and Price~\cite{storn_differential_1997}.
The results are shown in Fig.~\ref{fig:model_validation}\textit{B}, while the estimated parameter values are summarized in \ref{S1_Table}.
The parameter values are in very good agreement with order-of-magnitude values determined from the literature (\ref{S2_Text} and \ref{S1_Table}).

\section{Solution of optimal control problem}

The optimal control problem of Eq.~\ref{eq:J} consists in identifying the function $\alpha_{opt}(t)$ that maximizes the integral of the growth rate $\hat{\mu}$ over an interval $[0, \tau]$.
In order to solve this problem, we first redefined it over an infinite horizon (\textit{i.e.}, $\tau \rightarrow \infty $) in order to avoid boundary effects occurring over finite time intervals, in particular the depletion of precursors just before reaching $\tau$.
With $\mathcal{U}=\{\alpha:\mathbb{R}^+ \rightarrow [0,1] \}$ the set of admissible controls, the full optimization problem for the nondimensionalized system is given by
\begin{equation}\label{eq:meth_prob}
\max_{\alpha \in \mathcal{U}} J(\alpha)\equiv \int_0^{\infty} \hat{r}(\hat{t}) \, \dfrac{\hat{p}(\hat{t})}{K+\hat{p}(\hat{t})}\, d\hat{t}.
\end{equation}
Since $J(\alpha)$ diverges, we actually consider overtaking optimality:
A solution is overtaking optimal if its performance index catches up with the performance index of any other solution (\cite{carlson_infinite_1991}, see \ref{S3_Text} for details).

Necessary conditions on optimal trajectories can be obtained by the Infinite Horizon Maximum Principle \cite{carlson_infinite_1991}, an extension of the well-known Pontryagin Maximum Principle.
Analysis of the Hamiltonian of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim} and the associated adjoint system shows that the optimal trajectory is a concatenation of bang arcs ($\alpha(\cdot)=0$ or $\alpha(\cdot)=1$) and possibly a singular arc corresponding to the optimal steady state $(\hat{p}(t),\hat{r}(t))=(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$, that is, the steady state leading to the optimal growth rate $\hat{\mu}_{opt}^*$ in the new environment after the upshift (\ref{S3_Text}).
Moreover, from the Kelley condition~\cite{borisov_fullers_2000}, we can show that if the optimal trajectory has a singular arc, then it must enter this singular arc \textit{via} a chattering arc, \textit{i.e.}, with an infinite number of switches of $\alpha (\cdot)$ between 0 and 1 (\ref{S3_Text}).
The chattering arc is characterized by a switching curve $\hat{r}=\varphi(\hat{p})$ in the $(\hat{p},\hat{r})$-plane, which passes through $(\hat{p}_{opt}^*,\hat{r}_{opt}^*)$.
The switching curve divides the phase plane into two regions, such that $\alpha(t)$ switches to 0 when the system is in the region above $\varphi$ and to 1 when the system is below $\varphi$ (\ref{S3_Text} and Fig.~\ref{fig:optimalcontrol}).

The above results have led to the conjectured optimal solution of Eq.~\ref{eq:optcontrol}.
In parallel, we numerically solved the problem of Eq.~\ref{eq:meth_prob} by a direct method using the \texttt{bocop} software~\cite{bonnans_bocop_2012}.
A time discretization allows the optimal control problem to be transformed into a nonlinear optimization problem,
solved here by interior point techniques.
A discretization by a Lobatto IIIC formula (6th order) was used with 4000 time steps, and the relative tolerance for the NLP solver was set to $10^{-14}$.
The optimal trajectories thus obtained are composed of a chattering arc followed by a steady state corresponding to the singular arc (Fig.~\ref{fig:optimalcontrol}).
The switching curve $\varphi(\hat{p})$ was computed from numerical simulations with different initial conditions.

\section{Specification and analysis of control strategies}

As described in the \textit{Results} section, we are interested in control strategies satisfying the following conditions:
\begin{description}
\item[(C1)] The control laws are static functions of the system variables (as opposed to, for instance, functions that depend on derivatives or integrals of the variables).
\item[(C2)] For any given constant environment $E_M$, they drive the self-replicator system towards a unique stable steady state that is not trivial, \textit{i.e.}, with nonzero growth rate.
\item[(C3)] This steady state corresponds to the optimal steady state $(\hat{p}_{opt}^*, \hat{r}_{opt}^*)$, allowing growth at the maximal rate $\mu^*_{opt}$.
\end{description}
It can be directly verified from the functions $f$, $g$, and $h$ defining the nutrient-only, precursor-only, and on-off control strategies (Eqs~\ref{eq:f}, \ref{eq:precursorstrategy}, and \ref{eq:stratswitch}) that they are indeed static functions of the system variables (or the system input, in the case of the nutrient-only strategy). 
Here we show that the other two conditions are also satisfied for all three strategies.

Following Eq.~\ref{eq:f}, the nutrient-only strategy is defined by $\alpha = f(E_M)$, so that $\alpha$ is constant after the upshift.
As shown above and in \ref{S1_Text}, this means that the system controlled by the nutrient-only strategy has a single nontrivial stable steady state (Condition~C2). 
In addition, in this case the optimal steady state is attained for $\alpha^*_{opt}$ defined as in Eq.~\ref{eq:alpha_mu_optimal}, and the following function $f$ therefore guarantees Condition~C3:
\begin{equation}
\label{eq:meth_f}
f(E_M) = \frac{E_M + \sqrt{K E_M}}{E_M + 2\sqrt{K\, E_M} + 1}.
\end{equation}
In \ref{S1_Text}, it is shown that Eq.~\ref{eq:meth_f} is the only definition of $f$ satisfying all conditions.
\ref{S1_Fig} shows a plot of $f(E_M)$ together with a biologically plausible Michaelis-Menten approximation (Eq.~\ref{eq:MMApprox}).
